Towards a miniKanren with fair search strategies

### 2019-05-23 Paper/paper.tex
Too many changes to worry about.

\usepackage{listings}
\lstset{
  language=Scheme,
  basicstyle=\ttfamily,
  morekeywords={run,conde,run*,defrel,==,fresh,disj,conj,disj+},
  alsodigit={!\$\%&*+-./:<=>?@^_~},
  morecomment=[l]{;,\#{}lang},
  mathescape=true,
  literate={-inf}{{$^\infty$}}{1} {/fair}{{$_{fair}$}}{3}
  		   {→}{{$\rightarrow$}}{1}
  		   {×}{{$\times$}}{1}
  		   {==}{{$\equiv$}}{1}
  		   {conde}{{\textcolor[rgb]{0,.3,.7}{\textbf{cond$^e$}}}}{4}
  		   {repeato}{{repeat$^o$}}{6}
  		   {g0}{{g$_0$}}{2}
  		   {g1}{{g$_1$}}{2}
  		   {g2}{{g$_2$}}{2}
  		   {gl}{{g$_l$}}{2}
  		   {gr}{{g$_r$}}{2}
  		   {DFSi }{{DFS$_i$}}{3}
  		   {DFSbi}{{DFS$_{bi}$}}{3}
  		   {DFSf }{{DFS$_f$}}{3}
  		   {BFSser}{{BFS$_{ser}$}}{3}
}
%% scheme-list .

@@

miniKanren is a family of relational programming languages.
\citep{Friedman:2005:RS:1121583} \citet{friedman_reasoned_2018} introduce 
miniKanren and its
implementation in \emph{The Reason Schemer} 
and \emph{The Reasoned Schemer, 2nd Ed} (TRS2). \citet{byrd2017unified} have 
proved that miniKanren programs are useful in solving many problems as in 
miniKanren.org.

% miniKanren is a family of relational programming languages.
% \citet{friedman_reasoned_2018} introduce miniKanren and its implementation
% in \emph{The Reasoned Schemer, 2nd Ed} (TRS2). miniKanren programs, especially
% relational interpreters, have been proven to be useful in solving many problems 
% as in \citet{byrd2017unified}.

\NOTE[DPF]{I think that the abbreviations for the search strategies
should not come up until the Introduction.}
\NOTE[LKC]{Do you mean they should not come up in Abstract?}

\NOTE[DPF]{When we talk about complete and incomplete, I think we should
make it explicit that we are only talking about (run n (x y z) g ...) and
not (run* (x y z) g ...), but an alternative is that a stream is \emph{productive}
or \emph{unproductive}.}
\NOTE[LKC]{I have added a sentence at the place where we define
completeness.}

% \conde{} is the flagship disjunction operator in miniKanren.
A subtlety arises 
when a \conde{} contains many clauses: not every clause has an 
equal chance to contribute to the result. As an example, consider the following 
relation \texttt{repeato} and its invocation.

@@

The \conde{} in TR2, however, generates a less expected result.

@@

To summarize our contribution, we
\begin{itemize}
	\item propose and implement balanced interleaving depth-first search 
	(\DFSbi{}), a new search strategy with almost-fair \disj{}.
	\item propose and implement fair depth-first search (\DFSf{}), a 
	new search strategy with fair \disj{}.
	\item implement in a new way breath-first search (\BFSser), a search 
strategy with fair \disj{} and fair \conj{}. And we prove formally that our 
\BFSser{} implementation is semantically equivalent to \BFSser{}.
Our code runs faster in all benchmarks and we believe it is simpler.
\end{itemize}

@@

In this section, we define fair \disj{}, almost-fair \disj{} and fair \conj{}. 
Before going further into fairness, we give a short review of the terms:
\emph{state}, search \emph{space}, \emph{goal}.
A \emph{state} is a collection of constraints. (Here, we restrict 
constraints to unification constraints.) Every answer corresponds to a 
state. A space is a collection of states. And a \emph{goal} is a 
function from a state to a space.

@@

Given the following program, it is natural to expect lists of each letter to
constitute $1/4$ in the query result. \DFSi, the current search
strategy, however, results in many more lists of \texttt{a}s than lists
of other letters. And some letters  (e.g. \texttt{c} and \texttt{d}) are
rarely seen. The situation would be exacerbated if \conde{} would have
contained more clauses.

@@

\NOTE[LKC]{I might be wrong, but `\emph{so} evenly ... that the maximal ...' 
looks more natural to me}

@@

Given the following program, it is natural to expect lists of each letter to
constitute $1/4$ in the answer list. Search strategies with unfair \conj{} 
(e.g. 
\DFSi, \DFSbi, \DFSf), however, results in many more lists of \texttt{a}s than 
lists of other letters. And some letters are rarely seen. The situation would 
be exacerbated if \conde{} were to contain more clauses.
Although some strategies have different level of fairness in \disj{}, they 
have the same behavior when there is no call to a relational definition in 
\conde{} clauses (including this case).

@@

A more interesting situation is when the first conjunct produces an unbounded
number of states. Consider the following example, a naive specification of 
fair \conj{} 
might require search strategies to produce all sorts of singleton lists, but 
there
would not be any lists of length two or longer, which makes the strategies 
incomplete. 
A search strategy is \emph{complete} if and only if ``every correct answer 
would be discovered after some finite time'' \cite{seres1999algebra}, 
otherwise, it is \emph{incomplete}. In the 
context of miniKanren, a search strategy is complete means that every correct 
answer has a positition in large enough answer lists.

\NOTE[DPF]{Does this assume that we never use run*, since in fact, the paper 
does not include and uses of run*.  Something should be mentioned around here.}
\NOTE[LKC]{We never use run* in this paper because fairness is only 
interested when we have an unbounded number of answers. However, it is 
perfectly fine to use run* with any search strategies.}

@@

@@

\NOTE[DPF]{If we use the name subscript, the subscripts should use an 
underscore and we should do it for gunderscorel and gundescorer. If we can use 
an l that looks less like a 1, that would even be better.  Look at the l in 
TLT.}
\NOTE[LKC]{I have turned those `subscripts' to real subscripts.}

@@

\NOTE[DPF]{Why does the pair? case precede the null? case? Is it simply more 
efficient? It shouldn't be}
\NOTE[LKC]{I did for efficiency. I have re-ordered two cases back.}

@@

\NOTE[DPF]{When or where do you swap them back?}
\NOTE[LKC]{In the recursion}

@@

\subsection{optimized BFS}

@@

\NOTE[DPF]{Never refer to a literal section like 2.2. Instead
stick a label where it starts right near the section number and
stick another label where the subsection starts.
Then ref the two labels and place a period between them.}
\NOTE[LKC]{Got it! Fixed}

As mentioned in \autoref{sec:fairconj}, BFS puts answers in bags and 
answers of the 
same cost are in the same bag. The cost
information is recorded subtly -- the \texttt{car}s of a search space have cost 
0 (i.e. they are all in the same bag), and every thunk indicates an increment 
in cost. It is even more subtle that
\texttt{append$^\infty_{fair}$} and the \texttt{append-map$^\infty_{fair}$} 
respects the cost information. We make these facts more obvious by 
changing the type of search space, modifying related function definitions, 
and introducing a few more functions.

@@

We list functions related to the pure subset in Fig.~\ref{BFS-opt} (the 
others in Fig.~\ref{BFS-opt-cont}). We compare these functions with 
\citeauthor{seres1999algebra}'s implementation in our proof. The first three 
functions in Fig.~\ref{BFS-opt} are search space constructors. \texttt{none} 
makes an empty search space; \texttt{unit} makes a space from one answer; and 
\texttt{step} makes a space from a thunk. The remaining functions as before.

@@

\citet{kiselyov2005backtracking} has shown that a \emph{MonadPlus} hides in 
implementations of logic programming system. Our \BFSser{} implementation is 
not an exception: \texttt{append-map$^\infty_{fair}$} is like \texttt{bind}, 
but takes arguments in reversed order; \texttt{none}, \texttt{unit}, and 
\texttt{append$^\infty_{fair}$} correspond to \texttt{mzero}, \texttt{unit}, 
and \texttt{mplus} respectively.

\begin{figure}
	\lstinputlisting{Figures/BFS-opt-cont.rkt}	
	\caption{new and changed functions in optimized BFS that implements impure 
		features}
	\label{BFS-opt-cont}
\end{figure}

Functions implementing impure features are in Fig.~\ref{BFS-opt-cont}. The 
first function, \texttt{elim}, takes a space \texttt{s$^\infty$} and two 
continuations \texttt{ks} and \texttt{kf}. When \texttt{s$^\infty$} contains 
some 
answers, it calls \texttt{ks} with the first answer and the rest of the space. 
Otherwise, it calls \texttt{kf} with no argument. Here `s' and `f' means 
`succeed' and `fail' respectively. This function is similar to an eliminator of 
search spaces, hence the name. The remaining functions do the same thing as 
before.

@@

\NOTE[DPF]{Below, veryrecursiveo (if possible, should have its o raised.)
One way to do this is to have a different macro.  In the table there are
other occurrences that need their names with a different macro.}
\NOTE[LKC]{Fixed}

In general, only \DFSi{} and \DFSbi{} constantly perform well. \DFSf{} 
is just as 
efficient in all benchmarks but \veryrecursiveo. Both BFS have obvious 
overhead in many cases. Among the three variants of DFS (they all have unfair 
\conj{}), \DFSf{} is most resistant to clause permutation, followed by 
\DFSbi{} then 
\DFSi{}. Among the two implementation of BFS, ours constantly performs as well 
or 
better. Interestingly, every strategy with fair \disj{} suffers in 
\veryrecursiveo and \DFSf{} performs well elsewhere. 
Therefore, this 
benchmark might be a special case. Fair \conj{} imposes considerable overhead 
constantly except in \texttt{appendo}. The reason might be that strategies with 
fair \conj{} tend to keep more intermediate answers in the memory.

@@

\citet{yang2010adventures} points out a disjunct complex would be `fair' 
if it were a full and balanced tree.

@@

We devise two new search strategies (i.e. balanced interleaving DFS 
(\DFSbi{}) and fair DFS (\DFSf{})) and devise a new 
implementation of BFS. These strategies have different features 
in fairness: $_\textrm{bi}$ has an almost-fair \disj{} and unfair \conj{}. 
\DFSf{} has fair \disj{} and unfair \conj{}. \BFSser{} has both fair
\disj{} and fair \conj{}.

Our quantitative evaluation shows that \DFSbi{} and \DFSf{} are 
competitive 
alternatives to \DFSi{}, the current search strategy, and that 
\BFSser{} is less practical than other strategies.
\NOTE[DPF]{what?}
\NOTE[LKC]{other search strategies}

>>>>

\usepackage{listings}
\lstset{
  language=Scheme,
  basicstyle=\ttfamily,
  morekeywords={run,conde,run*,defrel,==,fresh,disj,conj,disj+},
  alsodigit={!\$\%&*+-./:<=>?@^_~},
  morecomment=[l]{;,\#{}lang},
  mathescape=true,
  literate={-inf}{{$^\infty$}}{1} {/fair}{{$_{fair}$}}{3}
  		   {→}{{$\rightarrow$}}{1}
  		   {×}{{$\times$}}{1}
  		   {==}{{$\equiv$}}{1}
  		   {conde}{{\textcolor[rgb]{0,.3,.7}{\textbf{cond$^e$}}}}{4}
                   {appendo}{{append$^o$}}{6}
  		   {repeato}{{repeat$^o$}}{6}
  		   {g0}{{g$_0$}}{2}
  		   {g1}{{g$_1$}}{2}
  		   {g2}{{g$_2$}}{2}
  		   {gl}{{g$_l$}}{2}
  		   {gr}{{g$_r$}}{2}
  		   {DFSi }{{DFS$_i$}}{3}
  		   {DFSbi}{{DFS$_{bi}$}}{3}
  		   {DFSf }{{DFS$_f$}}{3}
  		   {BFSser}{{BFS$_{ser}$}}{3}
}
%% scheme-list .

@@

miniKanren is a family of relational programming languages.
\citep{Friedman:2005:RS:1121583} \citet{friedman_reasoned_2018} introduce 
miniKanren and its
implementation in \emph{The Reasoned Schemer} 
and \emph{The Reasoned Schemer, 2nd Ed} (TRS2). \citet{byrd2017unified} have 
demonstrated that miniKanren programs are useful in solving several problems.
miniKanren.org contains the seeds of many more problems and their solutions.

\NOTE[DPF]{Yes, that is what I mean. And furthermore, I think that changing opt
to imp (for improved) makes more sense.  We may be optimizing, but it seems more
like an improvement.  I did change some of them from optimized to improved.
In my letter wrt to using
BFSopt or BFSimp, it still makes more sense to me.  Perhaps it should be part
of the title that we have improved on the the standard BFS implementation.}
\NOTE[LKC]{Do you mean they should not come up in Abstract?}
\NOTE[DPF]{When you rewrite the abstract and drop the abbreviations, remember when
you are rewriting the Introduction and introducing the abbreviations, that you should
explain the ser of BFSser using a reference.}

A subtlety arises 
when a \conde{} contains many clauses: not every clause has an 
equal chance of contributing to the result. As an example, consider the following 
relation \texttt{repeato} and its invocation.

@@

The \conde{} in TRS2, however, generates a less expected result.

@@

To summarize our contributions, we
\begin{itemize}
	\item propose and implement balanced interleaving depth-first search 
	(\DFSbi{}), a new search strategy with almost-fair \disj{}.
	\item propose and implement fair depth-first search (\DFSf{}), a 
	new search strategy with fair \disj{}.
	\item implement in a new way breath-first search (\BFSser), a search 
strategy with fair \disj{} and fair \conj{}. And we prove formally that our 
\BFSser{} implementation is semantically equivalent to \BFSser{}.
Our code runs faster in all benchmarks and we believe it is simpler.
\end{itemize}

@@

In this section, we define fair \disj{}, almost-fair \disj{} and fair \conj{}. 
Before going further into fairness, we give a short review of the terms:
\emph{state}, search \emph{space}, and \emph{goal}.
A \emph{state} is a collection of constraints. (Here, we restrict 
constraints to unification constraints.) Every answer corresponds to a 
state. A space is a collection of states. And a \emph{goal} is a 
function from a state to a space.

@@

Given the following program, it is natural to expect lists of each letter to
constitute $1/4$ in the query result. \DFSi, the current search
strategy, however, results in many more lists of \texttt{a}s than lists
of other letters. And some letters  (e.g. \texttt{c} and \texttt{d}) are
rarely seen. The situation would be exacerbated if the \conde{} would have
contained more clauses.

@@

@@

Given the following program, it is natural to expect lists of each letter to
constitute $1/4$ in the answer list. Search strategies with unfair \conj{} 
(e.g. 
\DFSi, \DFSbi, \DFSf), however, results in many more lists of \texttt{a}s than 
lists of other letters. And some letters are rarely seen. The situation would 
be exacerbated if \conde{} were to contain more clauses.
Although some strategies have a different level of fairness in \disj{}, they 
have the same behavior when there is no call to a relational definition in 
\conde{} clauses (including this case).

@@

A more interesting situation is when the first conjunct produces an unbounded
number of states. Consider the following example, a naive specification of 
fair \conj{} 
might require search strategies to produce all sorts of singleton lists, but 
there
would not be any lists of length two or longer, which makes the strategies 
incomplete. 
A search strategy is \emph{complete} if and only if ``every correct answer 
would be discovered after some finite time'' \cite{seres1999algebra}, 
otherwise, it is \emph{incomplete}. In the 
context of miniKanren, a search strategy is complete means that every correct 
answer has a position in large enough answer lists.

\NOTE[DPF]{I think that should be said as directly as in your note.}
\NOTE[LKC]{We never use run* in this paper because fairness is only 
interesting when we have an unbounded number of answers. However, it is 
perfectly fine to use run* with any search strategies.}

@@

\NOTE[DPF]{Next to the last line of the above paragraph the word same
has too many meanings.  It could be have the same cost or it could be
failing to distinguish two list that would be the same if we substituted
a b for an a.  It needs to be clarified.  It might help to work out
the actual numbers}

@@

\NOTE[DPF]{This came out fabulous.  I think I spotted some missed opportunities
later, which I will point out with additional notes.}
\NOTE[LKC]{I have turned those `subscripts' to real subscripts.}

@@

@@

\NOTE[DPF]{When or where do you swap them back? This could be still clearer.}
\NOTE[LKC]{In the recursion}

@@

\subsection{improved BFS}

@@

As mentioned in \autoref{sec:fairconj}, \BFSser{} puts answers in bags and 
answers of the 
same cost are in the same bag. The cost
information is recorded subtly---the \texttt{car}s of a search space have cost 
0 (i.e. they are all in the same bag), and every thunk indicates an increment 
in cost. It is even more subtle that
\texttt{append$^\infty_{fair}$} and the \texttt{append-map$^\infty_{fair}$} 
respects the cost information. We make these facts more obvious by 
changing the type of search space, modifying related function definitions, 
and introducing a few more functions.

@@

We list functions related to the pure subset in Fig.~\ref{BFS-opt} (the 
others in Fig.~\ref{BFS-opt-cont}). We compare these functions with 
\citeauthor{seres1999algebra}'s implementation in our proof. The first three 
functions in Fig.~\ref{BFS-opt} are search space constructors. \texttt{none} 
makes an empty search space; \texttt{unit} makes a space from one answer; and 
\texttt{step} makes a space from a thunk. The remaining functions are as before.

@@

\citet{kiselyov2005backtracking} have shown that a \emph{MonadPlus} hides in 
implementations of logic programming system. Our \BFSser{} implementation is 
not an exception: \texttt{append-map$^\infty_{fair}$} is like \texttt{bind}, 
but takes arguments in reversed order; \texttt{none}, \texttt{unit}, and 
\texttt{append$^\infty_{fair}$} correspond to \texttt{mzero}, \texttt{unit}, 
and \texttt{mplus}, respectively.

\begin{figure}
	\lstinputlisting{Figures/BFS-opt-cont.rkt}	
	\caption{new and changed functions in improved BFS that implement impure 
		features}
	\label{BFS-opt-cont}
\end{figure}

Functions implementing impure features are in Fig.~\ref{BFS-opt-cont}. The 
first function, \texttt{elim}, takes a space \texttt{s$^\infty$} and two 
continuations \texttt{ks} and \texttt{kf}. When \texttt{s$^\infty$} contains 
some 
answers, it calls \texttt{ks} with the first answer and the rest of the space. 
Otherwise, it calls \texttt{kf} with no argument. Here `s' means
`succeed' and `f' means `fail'. This function is similar to an eliminator of 
search spaces, hence the name. The remaining functions are as before.

@@

In general, only \DFSi{} and \DFSbi{} constantly perform well. \DFSf{} 
is just as 
efficient in all benchmarks but \veryrecursiveo. Both implementations of BFS have obvious 
overhead in many cases. Among the three variants of DFS, which all all have unfair \conj{}),
\DFSf{} is most resistant to clause permutation, followed by 
\DFSbi{} then 
\DFSi{}. Among the two implementation of BFS, our improved \BFSser{} constantly performs as well 
or 
better. Interestingly, every strategy with fair \disj{} suffers in 
\veryrecursiveo{} and \DFSf{} performs well elsewhere. 
Therefore, this 
benchmark might be a special case. Fair \conj{} imposes considerable overhead 
constantly except in \appendo. The reason might be that strategies with 
fair \conj{} tend to keep more intermediate answers in the memory.

\NOTE[DPF]{appendo is not raised, though I tried.}

@@

\citet{yang2010adventures} points out that a disjunct complex would be `fair' 
if it were a full and balanced tree.

@@

We devise two new search strategies (i.e. balanced interleaving DFS 
(\DFSbi{}) and fair DFS (\DFSf{})) and devise a new 
implementation of BFS. These strategies have different features 
in fairness: \DFSbi{} has an almost-fair \disj{} and unfair \conj{}. 
\DFSf{} has fair \disj{} and unfair \conj{}. \BFSser{} has both fair
\disj{} and fair \conj{}.

Our quantitative evaluation shows that \DFSbi{} and \DFSf{} are 
competitive 
alternatives to \DFSi{}, the current search strategy, and that 
\BFSser{} is less practical than other breadth-first strategies.
\NOTE[DPF]{I think what I have is okay, now.}
\NOTE[LKC]{other search strategies}
